### 2024-09-24 paper/paper.qmd
created first version of paper with only graphs and tables

---
title: "Are Lower Income Communities Given More Noise Exemptions"
subtitle: "My subtitle if needed"
author: 
  - Siddharth Gowda
thanks: "Open Data Toronto: https://open.toronto.ca/"
date: today
date-format: long
abstract: "TODO: FINISH THIS LATER"
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)

# TODO: switch area from kms2 to 10kms2
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

```{r}
#| label: tbl-fulldatasample
#| tbl-cap: "Sample of Noise Exmption Dataset"
#| echo: false
#| warning: false
#| message: false

display_column_names <- c("Permit Type", "Ward Number", "Exemption Issue Year",
                          "Expected Permit End Year", 
                          "Ward Total Area (km squared)",
                          "Ward Total Population", "Ward Median Income (CAD)")

noise_exemp_ward_data_og <- read_csv(
  "../data/analysis_data/clean_noise_exemp_ward_data.csv")

noise_exemp_ward_data_og %>% 
  select(!id) %>% 
  head(1) %>% 
  kable(col.names = display_column_names)
```

## Data History

```{r}
#| label: fig-permityears
#| tbl-cap: "TODO"
#| tbl-sub: ["TODO", "TODO", "TODO"]
#| layout-ncol: 2
#| echo: false
#| warning: false
#| message: false

noise_exemp_ward_data_og %>% group_by(issue_year) %>% 
  summarise(n = n_distinct(id)) %>% kable(
    col.names = c("Issue Year", "Number of Exemptions")
  )

noise_exemp_ward_data_og %>% group_by(expected_end_year) %>% 
  summarise(n = n_distinct(id)) %>% kable(
    col.names = c("Expected End Year", "Number of Exemptions")
  )

```

\newpage

## Types of Permits

```{r}
#| label: tbl-exemptioncategories
#| tbl-cap: "All Exemption Categories"
#| echo: false
#| warning: false
#| message: false

full_permit_type_table = noise_exemp_ward_data_og %>% group_by(permit_type) %>% 
  summarise(freq = n_distinct(id)) %>% 
  kable(col.names = c("Type", "Number of Permits"))

# Amplfied sound is for things like concerts or events

# Construction - self explanatory
# Continous pour (like pouring concerete, essentially like constructions)
# Large Carne (need carne to move things, similar to construction)
# Other sound - self explanatory

full_permit_type_table
```

```{r}
#| label: tbl-relevantexemptioncategories
#| tbl-cap: "Relevant Exemption Categories"
#| echo: false
#| warning: false
#| message: false

# Will merge these into construction, non-construction, other
noise_exemp_ward_data = noise_exemp_ward_data_og %>% 
  mutate(permit_type = case_when(
    str_detect(permit_type, "amplified") ~ "non-construction",
    str_detect(permit_type, "(construction|continuous|crane)") ~ "construction",
    str_detect(permit_type, "other") ~ "other",
    TRUE ~ "other"
  ))

construction_permit_type_table = noise_exemp_ward_data %>% 
  group_by(permit_type) %>% 
  summarise(freq = n_distinct(id)) %>% 
  kable(col.names = c("Type", "Number of Permits"))

construction_permit_type_table

```

\newpage

## Ward Information

```{r}
#| label: fig-wardtotalexemptions
#| fig-cap: "Noise Exemptions by Ward"
#| echo: false
#| warning: false
#| message: false
# ward, number of occurrences

ward_exemptions <- noise_exemp_ward_data %>% 
  select(ward, id, permit_type) %>% 
  group_by(ward) %>% 
  summarise(exemptions = n_distinct(id),
            construction_exemptions = sum(permit_type == "construction"),
            non_construction_exemptions = 
              sum(permit_type == "non-construction"),
            other_exemptions = sum(permit_type == "other"))

ward_exemptions_bar <- noise_exemp_ward_data %>% 
  ggplot(aes(x = ward, fill = permit_type)) +
  geom_bar() +
  theme_linedraw() +
  labs(x = "Ward Number", y = "Number of Noise Exemptions")

ward_exemptions_bar
```

```{r}
#| label: fig-warddetails
#| fig-cap: "All Relevant Ward Details"
#| fig-subcap: ["Ward Area","Ward Population", "Ward Median Income"]
#| layout-ncol: 3
#| echo: false
#| warning: false
#| message: false

ward_dataset <- noise_exemp_ward_data %>% 
  select(ward, ward_area_km2, ward_population, ward_median_income) %>%
  # remove duplicates that occur because of data being joined
  distinct()

ward_area_plot = ward_dataset %>% 
  ggplot(aes(x = ward, y = ward_area_km2)) +
  geom_col() +
  theme_linedraw() +
  labs(x = "Ward Number", y = "Area (km squared)")

ward_population_plot = ward_dataset %>% 
  ggplot(aes(x = ward, y = ward_population)) +
  geom_col() +
  theme_linedraw() +
  labs(x = "Ward Number", y = "Population")

ward_median_income_plot = ward_dataset %>% 
  ggplot(aes(x = ward, y = ward_median_income)) +
  geom_col() +
  theme_linedraw() +
  labs(x = "Ward Number", y = "Median Income (CAD)")

ward_area_plot
ward_population_plot
ward_median_income_plot
```

\newpage

## Income Noise Exemption Relationship

```{r}
#| label: fig-wardtotalexemptionsscatterplots
#| fig-cap: "Noise Exemptions by Ward"
#| fig-subcap: ["TODO", "TODO", "TODO"]
#| layout-ncol: 3
#| echo: false
#| warning: false
#| message: false

correlation_data <- noise_exemp_ward_data %>% 
  group_by(ward, ward_median_income, ward_area_km2, 
           ward_population) %>% 
  summarise(
    num_exemptions = n_distinct(id),
    construction_exemptions = sum(permit_type == "construction"),
    non_construction_exemptions = sum(permit_type == "non-construction"),
    other_exemptions = sum(permit_type == "other")
  )

income_num_exemption <- correlation_data %>% 
  ggplot(aes(ward_median_income, num_exemptions)) +
  geom_point(alpha = 0.7) +
  theme_linedraw() +
  labs(x = "Median Income", y = "Number of Exemptions") + 
  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)

income_nonconstruction_exemptions <- correlation_data %>% 
  ggplot(aes(ward_median_income, non_construction_exemptions)) +
  geom_point(alpha = 0.7) +
  theme_linedraw() +
  labs(x = "Median Income", y = "Number of Non-construction Exemptions") + 
  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)

income_construction_exemptions <- correlation_data %>% 
  ggplot(aes(ward_median_income, construction_exemptions)) +
  geom_point(alpha = 0.7) +
  theme_linedraw() +
  labs(x = "Median Income", y = "Number of Construction Exemptions") + 
  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)

income_num_exemption
income_nonconstruction_exemptions
income_construction_exemptions

# TODO: delete before submission
test = correlation_data %>% filter(ward != 10)
cor.test(test$ward_median_income, test$num_exemptions)
cor.test(correlation_data$ward_median_income, correlation_data$num_exemptions)

# keep in mind that relationship looks weaker after potetnial outlier point
# removed

# TODO: ADD a section about these graphs even after they have been removed
```

\newpage

## Testing for Hidden Confounding relationships

```{r}
#| label: fig-confoudingplots
#| fig-cap: "Ward Median Income Relationship With Common Potential Confounds"
#| fig-subcap: ["TODO", "TODO"]
#| layout-ncol: 3
#| echo: false
#| warning: false
#| message: false

correlation_data <- noise_exemp_ward_data %>% 
  select(ward, ward_area_km2, ward_population, ward_median_income)

ward_income_pop <- correlation_data %>% filter(ward != 10) %>% 
  ggplot(aes(ward_median_income, ward_population)) +
  geom_point(alpha = 0.7) +
  theme_linedraw() +
  labs(x = "Median Income", y = "Population") + 
  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)

ward_income_area <- correlation_data %>% 
  ggplot(aes(ward_median_income, ward_area_km2)) +
  geom_point(alpha = 0.7) +
  theme_linedraw() +
  labs(x = "Median Income", y = "ward_area_plot (kms squared)") + 
  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)

ward_income_pop
ward_income_area

# TODO: DELETE BELOW CODE

cor.test(correlation_data$ward_median_income, correlation_data$ward_population)
cor.test(correlation_data$ward_median_income, correlation_data$ward_area_km2)

test = correlation_data %>% filter(ward != 10)
cor.test(test$ward_median_income, test$ward_population)
cor.test(test$ward_median_income, test$ward_area_km2)

# keep in mind that relationship looks weaker after potetnial outlier point
# removed (this is especially true for population)
```

```

\newpage

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses/Limitations and next steps should also be included.

\newpage

# References

### 2024-09-24 paper/paper.qmd
finished all writing first draft except for Discussion

@@
 date-format: long
-abstract: "TODO: FINISH THIS LATER"
+abstract: "TODO: FINISH THIS LATER (ABSTRACT)"
 format: pdf
@@
 # TODO: switch area from kms2 to 10kms2
+# TODO: Add a table of contents
+# TODO: adding planning table and graph sketches to other
+# TODO add bibliography properly
+# TODO: add a brief section in the intro talking about the different types of
+# TODO: add section citations
+# of permit
 ```
+
+# Table of Contents
+
+TODO: Work in Progress
 
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+Noise pollution is a overlooked, but serious and common problem in modern cities.
+Noise pollution has been linked with heart issues, irregular sleep, hearing loss, and mental health issues (Iberdrola 2024 - fix citation later).
+Hence, Toronto city officials have identified noise pollution as a problem and created @NoiseBylaws to limit at what times certain noises like construction can occur.
+However, sometimes noise needs to be allowed during quiet hours.
+For instance, some building need to continue pouring cement overnight, otherwise the building will have safety risks.
+As a result, the Toronto government has also created a system of noise exemptions.
 
-The remainder of this paper is structured as follows. @sec-data....
+This paper explores whether or not these exemptions are getting abuse.
+Specifically, we want to know if lower-income communities are more likely to get noise exemptions.
+This is part of a much larger phenomenon, which is that poorer communities are suffer more from noise pollution compared to affluent ones.
+[Insert sentence showing the results from some external study] Moreover, all this does is further increase lifestyle differences between low-income and higher income-individuals.
+So, it is critical to make sure that Toronto's noise exemptions are not furthering this gap, especially since noise exemptions are often glance over in terms of their addition to noise pollution compared to other contributors like late night traffic and airport plane landings.
+This paper attempts to fill that gap.
+
+To do this, we used public available data on noise exemptions provided by the government of Toronto and we mapped the area of the exemption to the area's income.
+This was done by estimating the income by looking at the ward where the permit's company is working in.
+From that, we determined if there was a relationship between the two variables.
+Furthermore, we look at other factors such the type of permits given when determining the relationship.
+Overall, we determined that there was no evidence that lower-income communities were more likely to have more noise exemptions.
+On the contrary, we found moderate evidence for the opposite: higher income communities had more noise exemptions, not matter the type of exemptions that were given.
+This is important because it provides some evidence that noise exemptions are not disproportionately creating more noise pollution for lower income communities.
+
+In terms of the paper's structure, it is organized into three main sections not including the introduction and table of Contents.
+The section # explores the data, its origins, potential issues, and relationships, while the section # provides a deeper discussion of the findings from the data.
+Finally, the section # addresses the paper's limitations and suggests potential next steps for future research.
 
 # Data {#sec-data}
+
+## Overview
+
+All data for this investigation was provided by @OpenDataToronto.
+This data sets after @sec-datacleaning came to 1227 noise exemption permit details.
+Furthermore, this @OpenDataToronto library is created and maintained by government of Toronto.
+Specifically, this data was obtained from three different data sets provided in the library.
+The first data set was obtained from the Noise Exemption Permits package.
+This data set included everything related to noise exemption permits, including, the day the permit was given, the permit type, the conditions behind it's use, the ward the permit is operating in and more.
+More about this in @tbl-fulldatasample.
+This data set was last updated 2024-09-23.
+
+The rest of the data sets come from the Ward Profiles (25-Ward Model).
+This package contains all the information someone would need about Toronto ward's.
+It was last updated 2024-09-17.
+As for the data sets, the more important one comes from the 2021 Census.
+It contains demographic and socioeconomic data about all of the wards in Toronto.
+It's important to note here that some of the data in that dataset was created based on 25% samples of the census.
+Consequently, variable that were created based on 25% census samples will be denoted in @tbl-fulldatasample The final data set is simply a mapping of the each ward number to the ward's total area in squared kilometers.
+
+All data analysis was performed using @citeR, @tidyverse, @janitor, and @ggplot2. Also, @kintir was used in order to finalize and create the output of this report.
+
+## Data Cleaning {#sec-datacleaning}
+
+Some noise exemption data had missing values and those were rows.
+The main reason this occurred was because quite a few of noise exemption permit entries did not have ward numbers.
+Also all permit issue date and permit expected end date values were all converted to years for simplicity purposes as the paper is not concerned about exactly when these permits were issued and ended.
+It's is only concerned about when these permits were issued to determine whether the data is relevant to present day Toronto.
+Additionally, all three of data sets were merged into one data set so each permit has all the information is needs about the ward it is residing in.
+This was soley done to make creating graphs, tables, and other forms of analysis easier.
 
@@
 
-## Data History
+| Variable Name           | Description                                                   |
+|-------------------------|---------------------------------------------------------------|
+| Permit Type             | The type of the permit. More on this in @sec-permittypes       |
+|-------------------------|---------------------------------------------------------------|
+| Ward Number             | The number of the ward of the address the permit's company is working in  |
+|-------------------------|---------------------------------------------------------------|
+| Exemption Issue Year     | The year the permit was issued                                |
+|-------------------------|---------------------------------------------------------------|
+| Expected Permit End Year | The year the government expects this permit to end            |
+|-------------------------|---------------------------------------------------------------|
+| Ward Total Area          | The total area the ward takes in squared kilometers           |
+|-------------------------|---------------------------------------------------------------|
+| Ward Population          | The number of people that live in the ward                   |
+|-------------------------|---------------------------------------------------------------|
+| Ward Median Income       | The median income of people in that ward (CAD)               |
+
+### Data Years
 
@@
 #| tbl-cap: "TODO"
-#| tbl-sub: ["TODO", "TODO", "TODO"]
+#| tbl-sub: ["TODO", "TODO"]
 #| layout-ncol: 2
@@
 
+The table shows that almost all permits have been issued in 2024 and expected to end sometime in 2024 or 2025.
+It's is important to note that this was not deliberate choice and that this is merely a limitation of the data.
+However, this should not be viewed as a limitation because having all of the data being from 2019 and beyond makes sure that all this data is relevant to how noise exemption permits are issued now.
+
 \newpage
 
-## Types of Permits
+### Types of Permits {#sec-permittypes}
 
@@
 ```
+
+Noise permits are applied to for a variety of reasons.
+These are all the ones listed in the data set.
+In reality there are more and take a look @NoiseToronto if interested in finding out more.
+It's clear that amplified sounds which are often required for things like concerted are the most common form of noise permit exemption.
+
+| Permit Type           | Description                                                   |
+|-------------------------|---------------------------------------------------------------|
+| Amplified Sound         | Sound made by an electric device. Often required for musical festivals.|
+|-------------------------|---------------------------------------------------------------|
+| Construction             | Self explanatory (TODO)  |
+|-------------------------|---------------------------------------------------------------|
+| Continuous Pour     | Continuously pouring some liquid, like concrete, that cannot be interrupted|
+|-------------------------|---------------------------------------------------------------|
+| Large Crane | Some sort of work that requires a crane (usually for a high rise building)|
+|-------------------------|---------------------------------------------------------------|
+| Other Sound          | Self-explanatory           |
+|-------------------------|---------------------------------------------------------------|
+
+TODO (will add the rest of the types later).
+
+However, for this analysis we will group the permit types into construction related, non-constructed related, and other.
+This is done not only to make things simpler, but because often times these construction related permit types are necessary and cannot be avoid and could potentially benefit a community while things like amplified sound are usually not necessary.
+
+Non-construction type includes: Amplified
+
+Construction type includes: Construction, Continuous Pour, Large Crane
+
+Other includes: Other Sounds
 
@@
 
-\newpage
+Even after merging the construction types together. Non-construction or amplified sound types are by far the most common type permit, almost taking up 75% of permits.
 
-## Ward Information
+## Exemption Distribution by Ward
 
@@
 #| message: false
+
+ward_exemptions_bar <- noise_exemp_ward_data %>% 
+  ggplot(aes(x = ward, fill = permit_type)) +
+  geom_bar() +
+  theme_linedraw() +
+  labs(x = "Ward Number", y = "Number of Noise Exemptions")
+
+# TODO: add summary statistics for this graph
+
+ward_exemptions_bar
+```
+
+```{r}
+#| label: tbl-wardexemptionssummary
+#| tab-cap: "Noise Exemptions Summary And Outliers"
+#| fig-subcap: ["Summary", "Outliers"]
+#| layout-ncol: 2
+#| echo: false
+#| warning: false
+#| message: false
+
 # ward, number of occurrences
-
 ward_exemptions <- noise_exemp_ward_data %>% 
@@
 
-ward_exemptions_bar <- noise_exemp_ward_data %>% 
-  ggplot(aes(x = ward, fill = permit_type)) +
-  geom_bar() +
-  theme_linedraw() +
-  labs(x = "Ward Number", y = "Number of Noise Exemptions")
+five_number_summary <- ward_exemptions %>%
+  summarise(
+    min = min(exemptions),
+    Q1 = quantile(exemptions, 0.25),
+    median = median(exemptions),
+    Q3 = quantile(exemptions, 0.75),
+    max = max(exemptions),
+    IQR = IQR(exemptions)
+  )
 
-ward_exemptions_bar
+# calculate outliers
+# TODO: add titles
+ward_exemption_outliers <- ward_exemptions %>% filter(
+  exemptions < five_number_summary$Q1 - 1.5*five_number_summary$IQR |
+    exemptions > five_number_summary$Q3 + five_number_summary$IQR) %>% 
+  select(ward, exemptions)
+
+five_number_summary %>% kable()
+ward_exemption_outliers %>% kable()
 ```
+Based on the @fig-wardtotalexemptions we can see that most wards give out more non-construction permits that construction permits.
+Hence, it's important to see how income is related to number of exemptions based on each type, so general data results are not biased towards non-constructive exemptions.
 
-```{r}
-#| label: fig-warddetails
-#| fig-cap: "All Relevant Ward Details"
-#| fig-subcap: ["Ward Area","Ward Population", "Ward Median Income"]
-#| layout-ncol: 3
-#| echo: false
-#| warning: false
-#| message: false
-
-ward_dataset <- noise_exemp_ward_data %>% 
-  select(ward, ward_area_km2, ward_population, ward_median_income) %>%
-  # remove duplicates that occur because of data being joined
-  distinct()
-
-ward_area_plot = ward_dataset %>% 
-  ggplot(aes(x = ward, y = ward_area_km2)) +
-  geom_col() +
-  theme_linedraw() +
-  labs(x = "Ward Number", y = "Area (km squared)")
-
-ward_population_plot = ward_dataset %>% 
-  ggplot(aes(x = ward, y = ward_population)) +
-  geom_col() +
-  theme_linedraw() +
-  labs(x = "Ward Number", y = "Population")
-
-ward_median_income_plot = ward_dataset %>% 
-  ggplot(aes(x = ward, y = ward_median_income)) +
-  geom_col() +
-  theme_linedraw() +
-  labs(x = "Ward Number", y = "Median Income (CAD)")
-
-ward_area_plot
-ward_population_plot
-ward_median_income_plot
-```
+Moreover, from @fig-wardtotalexemptions and @tbl-wardexemptionssummary, it's clear that wards 13, 11, and especially ten are outlines in terms of number of exemptions given compared to the rest of the data.
+However, for the purpose of this paper, we will not consider these data points as outliers, so they will not be discarded from the analysis.
+This is these are still wards of Toronto and their wards cannot be excluded when determining any conclusion or relationships.
 
@@
 #| message: false
+
+# TODO: change this so that the general num_exemptions graph is
+# in it's own code black and the sub groups have a layout-ncol-2
 
@@
 
-# TODO: delete before submission
-test = correlation_data %>% filter(ward != 10)
-cor.test(test$ward_median_income, test$num_exemptions)
-cor.test(correlation_data$ward_median_income, correlation_data$num_exemptions)
-
 # keep in mind that relationship looks weaker after potetnial outlier point
@@
 
-\newpage
+Based on the @fig-wardtotalexemptionsscatterplots.
+There is not seem to be a negative relation ship between median ward income and number of exemptions.
+However, by looking at the plot, there does seem to me a moderate postive relationship between median income and number of exemptions for all types of permit, although the constrution permit plot's relationship does apear to be weaker.
+
+It is important to note that if we hypothetically removed the outliers (ward 10, 11, 13), the relationship would appear to be a bit weaker, but the relationship still seems reasonable.
 
@@
 #| fig-subcap: ["TODO", "TODO"]
-#| layout-ncol: 3
+#| layout-ncol: 2
 #| echo: false
@@
 
-ward_income_pop <- correlation_data %>% filter(ward != 10) %>% 
-  ggplot(aes(ward_median_income, ward_population)) +
+ward_income_pop <- correlation_data %>% 
+  ggplot(aes(ward_population, ward_median_income, )) +
   geom_point(alpha = 0.7) +
   theme_linedraw() +
-  labs(x = "Median Income", y = "Population") + 
+  labs(y = "Median Income", x = "Population") + 
   geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)
@@
 ward_income_area <- correlation_data %>% 
-  ggplot(aes(ward_median_income, ward_area_km2)) +
+  ggplot(aes(ward_area_km2, ward_median_income)) +
   geom_point(alpha = 0.7) +
   theme_linedraw() +
-  labs(x = "Median Income", y = "ward_area_plot (kms squared)") + 
+  labs(y = "Median Income", x = "ward area (kms squared)") + 
   geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)
@@
 ward_income_area
+```
+@fig-confoudingplots explores if median income can be predicted by a ward's population size or area.
+Based on the plots, there does not appear to be a reltionship between ward median income and the ward's area.
+As for population, there does seem to be a moderate to weak relationship between ward population and median income.
 
-# TODO: DELETE BELOW CODE
-
-cor.test(correlation_data$ward_median_income, correlation_data$ward_population)
-cor.test(correlation_data$ward_median_income, correlation_data$ward_area_km2)
-
-test = correlation_data %>% filter(ward != 10)
-cor.test(test$ward_median_income, test$ward_population)
-cor.test(test$ward_median_income, test$ward_area_km2)
-
-# keep in mind that relationship looks weaker after potetnial outlier point
-# removed (this is especially true for population)
-```
-
-```
+Nevertheless, if we hypotectically remove the outliers from population graph. Then there is no relationship between population and median income.
 
@@
 
-## Weaknesses and next steps
+## Weaknesses and Next Steps
 
@@
 
+<!-- TODO: DELETE THIS CODE AND MOVE IT SOMEWHERE ELSE  -->
+
+# Special Thanks
+
+@citeR and @rohan.
+
+\newpage
+
 # References

### 2024-09-24 scripts/00-simulate_data.R

@@ -31,7 +31,7 @@ sim_noise_exmp_data <- tibble(
   type = sample(permit_types, size = n, replace = TRUE),
   ward = sample(wards, size = n, replace = TRUE),
   issue_year = sample(issue_years, size = n, replace = TRUE),
-  poverty_rate = rnorm(n, 0.10, 0.03),
+  median_income = rnorm(n, 50000, 13000),
   population = floor(rnorm(n, 100000, 25000)),
   ward_area_kms_sq = rnorm(n, 30, 10)
 )
@@ -46,7 +46,7 @@ sim_noise_exmp_data <- sim_noise_exmp_data %>%
   mutate(id = row_number())
 
 write_csv(sim_noise_exmp_data, 
-         "./data/simulation_data/sim_noise_exmp_data.csv")
+         "./data/simulation_data/sim_noise_exemp_data.csv")
 
 
 

### 2024-09-24 scripts/03-test_data.R

@@ -15,30 +15,30 @@ library(tidyverse)
 #### Test data ####
 
 ## simulated data tests
-sim_noise_exmp_data <- 
-  read_csv("./data/simulation_data/sim_noise_exmp_data.csv")
+sim_noise_exemp_data <- 
+  read_csv("./data/simulation_data/sim_noise_exemp_data.csv")
 
 print("Testing No Na's\n")
-count(sim_noise_exmp_data %>% drop_na()) == count(sim_noise_exmp_data)
+count(sim_noise_exemp_data %>% drop_na()) == count(sim_noise_exemp_data)
 
-print("Testing Poverty Rate is a percentage")
-count (sim_noise_exmp_data %>% 
-         filter(poverty_rate > 1 | poverty_rate < 0)) == 0
+print("Testing only 25 wards from 1-25")
+count (sim_noise_exemp_data %>% 
+         filter(!(ward %in% c(1:25)))) == 0
 
 print("Issue Year always the same or less than the expected end year")
-count (sim_noise_exmp_data %>% 
+count (sim_noise_exemp_data %>% 
          filter(issue_year > expected_end_year)) == 0
 
 ## cleaned data tests
 noise_exmp_ward_data <- read_csv(
-  "./data/analysis_data/clean_noise_exmp_ward_data.csv")
+  "./data/analysis_data/clean_noise_exemp_ward_data.csv")
 
 print("Testing No Na's\n")
 count(noise_exmp_ward_data %>% drop_na()) == count(noise_exmp_ward_data)
 
-print("Testing Poverty percentage is a percentage")
+print("Testing only 25 wards from 1-25")
 count (noise_exmp_ward_data %>% 
-         filter(poverty_rate > 1 | poverty_rate < 0)) == 0
+         filter(!(ward %in% c(1:25)))) == 0
 
 print("Issue Year always the same or less than the expected end year")
 count (noise_exmp_ward_data %>% 

### 2024-09-24 paper/paper.qmd
fixed confound section

@@
 
-\newpage
-
 ### Types of Permits {#sec-permittypes}
@@
 It's clear that amplified sounds which are often required for things like concerted are the most common form of noise permit exemption.
+
+TODO: add name and captions
 
@@
 #| tab-cap: "Noise Exemptions Summary And Outliers"
-#| fig-subcap: ["Summary", "Outliers"]
+#| tbl-subcap: ["Summary", "Outliers"]
 #| layout-ncol: 2
@@
 This is these are still wards of Toronto and their wards cannot be excluded when determining any conclusion or relationships.
-
-\newpage
 
@@
 There is not seem to be a negative relation ship between median ward income and number of exemptions.
-However, by looking at the plot, there does seem to me a moderate postive relationship between median income and number of exemptions for all types of permit, although the constrution permit plot's relationship does apear to be weaker.
+However, by looking at the plot, there does seem to me a moderate positive relationship between median income and number of exemptions for all types of permit, although the construction permit plot's relationship does appear to be weaker.
 
-It is important to note that if we hypothetically removed the outliers (ward 10, 11, 13), the relationship would appear to be a bit weaker, but the relationship still seems reasonable.
+It is important to note that if we hypothetically removed the outliers (ward 10, 11, 13), the relationship would appear to be a bit weaker, but there still appears to be a moderate positive relationship.
 
@@
 #| label: fig-confoudingplots
-#| fig-cap: "Ward Median Income Relationship With Common Potential Confounds"
+#| fig-cap: "Ward Population impact on Medium Impact"
 #| fig-subcap: ["TODO", "TODO"]
@@
 
-correlation_data <- noise_exemp_ward_data %>% 
+pop_exemptions = correlation_data %>% ggplot(aes(ward_population, num_exemptions)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(y = "Exemptions", x = "Population") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)
+
+correlation_data_confound <- noise_exemp_ward_data %>% 
   select(ward, ward_area_km2, ward_population, ward_median_income)
 
-ward_income_pop <- correlation_data %>% 
-  ggplot(aes(ward_population, ward_median_income, )) +
+ward_income_pop <- correlation_data_confound %>% 
+  ggplot(aes(ward_population, ward_median_income)) +
   geom_point(alpha = 0.7) +
@@
 
-ward_income_area <- correlation_data %>% 
-  ggplot(aes(ward_area_km2, ward_median_income)) +
-  geom_point(alpha = 0.7) +
-  theme_linedraw() +
-  labs(y = "Median Income", x = "ward area (kms squared)") + 
-  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)
+# ward_income_area <- correlation_data %>% 
+#   ggplot(aes(ward_area_km2, ward_median_income)) +
+#   geom_point(alpha = 0.7) +
+#   theme_linedraw() +
+#   labs(y = "Median Income", x = "ward area (kms squared)") + 
+#   geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)
 
 ward_income_pop
-ward_income_area
+pop_exemptions
+# ward_income_area
+
 ```
-@fig-confoudingplots explores if median income can be predicted by a ward's population size or area.
-Based on the plots, there does not appear to be a reltionship between ward median income and the ward's area.
-As for population, there does seem to be a moderate to weak relationship between ward population and median income.
+@fig-confoudingplots explores a ward's population is a good predictor of a ward's population and the number of permit exemptions given.
+Based on the plots there does seem to be a moderate positive relationship between a ward's population and a ward's medium income.
+Likewise, there is a moderate positive relationship between ward population and the number of noise exemptions a ward has given although this reltionship is not as strong as population and median income.
 
-Nevertheless, if we hypotectically remove the outliers from population graph. Then there is no relationship between population and median income.
+However, we need to also consider that if we hypothetically remove the outlines from population graph.
+Then there appeared to be very little if not no relationship between population and median income.
 

### 2024-09-24 paper/paper.qmd
first full draft of discussion

@@
 
-## Income Noise Exemption Relationship
+## Income Noise Exemption Relationship {#sec-relationships}
 
@@
 Based on the plots there does seem to be a moderate positive relationship between a ward's population and a ward's medium income.
-Likewise, there is a moderate positive relationship between ward population and the number of noise exemptions a ward has given although this reltionship is not as strong as population and median income.
+Likewise, there is a moderate positive relationship between ward population and the number of noise exemptions a ward has given although this relationship is not as strong as population and median income.
 
@@
 
-## First discussion point {#sec-first-point}
+TODO: add correlation numbers, and other numbers
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+Based on @sec-relationships, we have evidence that suggests that noise exemption permits are not disproportionately given out to lower-income communities.
+This was done by look at the relationship between medium income of wards and the total permits given in those wards.
+Furthermore, after cases for the types of permits given, we found no evidence for their being more construction related permits or non-constructions permits given to lower income wards.
 
-## Second discussion point
+This finding can be used in a variety of ways.
+For one, since we know that usually lower income communities are more likely to affected by noise pollution, we now understand that this noise pollution typically does not come from government noise exemption permits.
+Hence, city staff and officials should focus their efforts on combating other forms of noise pollution in low-income eras.
 
-## Third discussion point
+## Unexpected Relationship {#sec-first-point}
+
+Surprisingly, we discovered that there is a moderate positive relationship between a ward's median income and the number of exemptions given.
+This was also true when casing for construction and non-construction based permits.
+For the construction permits, it is reasonable to assume that higher income-areas have more construction permits since more companies will want to invest their, build building, and those same companies and people will need to continue to maintain those building.
+So, the city could potentially benefit by investing more in these lower income areas.
+However, it is a bit unclear why there are more non-construction (amplified sound) exemptions.
+Nevertheless, city officials could use this finding to discuss whether or not these exemptions for amplified sound are being given out too frequently especially since they are oftentimes not necessary for growing the community like construction related ones are.
+
+## Potential Confounds
+
+A ward's population was also found to be associated with a ward's number of noise exemptions.
+The relationship is understandable because ward's with more people will have more activities and more construction to meet the needs of those people, so it makes sense to have those permits.
 
@@
 
-Weaknesses/Limitations and next steps should also be included.
+The biggest weakness in this study was the use of wards as aggregators.
+The loss of specificity of the data when using wards made a few outliers (Wards 10, 11, and 13) have significant impact on the potential relationships that could be formed with income and noise exemptions.
+Unfortunate, wards 10, 11, and 13 cannot be ignored until most outliers.
+This is because there are only 25 wards in Toronto, so removing three data points is removing a lot of data.
+Moreover though, the data from these wards are not junk they are accurate reflection how noise and income are related in Toronto.
 
-\newpage
+Additionally, using wards to determine socio-economic status can quickly lead to over generalized results.
+In particular, simply because a ward on average is richer than other wards, it does not mean that particular area where the noise permit has been used given is a rich area.
 
-<!-- TODO: DELETE THIS CODE AND MOVE IT SOMEWHERE ELSE  -->
-
-# Special Thanks
-
-@citeR and @rohan.
+In terms of next steps, the first step would be to determine even more specifically the socioeconomic status of an area based on a specific address given in the application permit.
+Then based on this do similar analysis to determine if there is a true relationship between income and noise permits.
+Furthermore, in future studies regarding this topic, we would benefit from trying to model the relationship income and noise exemptions.
+Not only does create a more in depth quantification of the relationship, it also provides a much more throughout way of factoring for confounding variables.
 

### 2024-09-24 paper/paper.qmd
edited dicussion

@@
 The relationship is understandable because ward's with more people will have more activities and more construction to meet the needs of those people, so it makes sense to have those permits.
+This might invalidate the finding that more median income has a positive relationship with number of permits for a ward if there is a relationship with population and ward medium income: and there is a moderate positive relationship.
+However, after removing the potential outliers (ward 10, 11, and 13) there appears to me no relationship between the two variables.
+Therefore, further explore would need to be done to fix determine whether or no this is a confounding relationship.
+See section @sec-weaknessnextsteps for more dicussion on confounding variables.
 
-## Weaknesses and Next Steps
+## Weaknesses and Next Steps {#sec-weaknessnextsteps}
 

### 2024-09-26 paper/paper.qmd
reran entire repo to test repoducability and removed starter files

--- 2024-09-24 paper/paper.qmd
@@
 ---
-title: "Are Lower Income Communities Given More Noise Exemptions"
-subtitle: "My subtitle if needed"
+title: "Lower Income And Minority Communities Are Not Given More Noise Exemptions"
 author: 
   - Siddharth Gowda
-thanks: "Open Data Toronto: https://open.toronto.ca/"
 date: today
 date-format: long
-abstract: "TODO: FINISH THIS LATER (ABSTRACT)"
+thanks: "Code and data are available at: https://github.com/siddharthgowda/Toronto_Noise_Exmptions_Income "
+abstract: "Low-income and minority communities often incur the worst of noise pollution which in turn lowers their quality of life.
+This paper investigates if low-income and minority communities are more likely to receive noise permit exemptions.
+Contrary to expectations, the findings indicate that low-income and minority communities do not have a higher likelihood of receiving exemptions, In fact we found evidence for the opposite being true, that communities with higher income and less minorities recieved more noise exemption permits."
 format: pdf
@@
 library(knitr)
-
-# TODO: switch area from kms2 to 10kms2
-# TODO: Add a table of contents
-# TODO: adding planning table and graph sketches to other
-# TODO add bibliography properly
-# TODO: add a brief section in the intro talking about the different types of
-# TODO: add section citations
-# of permit
+library(scales)
 ```
-
-# Table of Contents
-
-TODO: Work in Progress
 
@@
 Noise pollution is a overlooked, but serious and common problem in modern cities.
-Noise pollution has been linked with heart issues, irregular sleep, hearing loss, and mental health issues (Iberdrola 2024 - fix citation later).
-Hence, Toronto city officials have identified noise pollution as a problem and created @NoiseBylaws to limit at what times certain noises like construction can occur.
+Noise pollution has been linked with heart issues, irregular sleep, hearing loss, and mental health issues [@IberdrolaNoisePollution].
+Hence, Toronto city officials have identified noise pollution as a problem and created Noise Bylaws [@torontonoise] to limit at what times certain noises like construction can occur.
 However, sometimes noise needs to be allowed during quiet hours.
@@
 This paper explores whether or not these exemptions are getting abuse.
-Specifically, we want to know if lower-income communities are more likely to get noise exemptions.
-This is part of a much larger phenomenon, which is that poorer communities are suffer more from noise pollution compared to affluent ones.
-[Insert sentence showing the results from some external study] Moreover, all this does is further increase lifestyle differences between low-income and higher income-individuals.
-So, it is critical to make sure that Toronto's noise exemptions are not furthering this gap, especially since noise exemptions are often glance over in terms of their addition to noise pollution compared to other contributors like late night traffic and airport plane landings.
-This paper attempts to fill that gap.
+Specifically, we want to know if lower-income and minority communities are more likely to get noise exemptions.
+This is part of a much larger phenomenon, which is that poorer and minority communities suffer more from noise pollution compared to affluent ones [@Seltenrich2018].
+Moreover, all this does is further increase disadvantages low-income communities and minorities have.
+So, it is critical to make sure that Toronto's noise exemptions are not further dividing this gap, especially since noise exemptions are often glanced over in terms of their addition to noise pollution compared to other contributors like late night traffic and airport plane landings.
 
-To do this, we used public available data on noise exemptions provided by the government of Toronto and we mapped the area of the exemption to the area's income.
+To explore this topic, we used public available data on noise exemptions provided by the government of Toronto and we mapped the area of the exemption to the area's income.
 This was done by estimating the income by looking at the ward where the permit's company is working in.
-From that, we determined if there was a relationship between the two variables.
-Furthermore, we look at other factors such the type of permits given when determining the relationship.
-Overall, we determined that there was no evidence that lower-income communities were more likely to have more noise exemptions.
-On the contrary, we found moderate evidence for the opposite: higher income communities had more noise exemptions, not matter the type of exemptions that were given.
-This is important because it provides some evidence that noise exemptions are not disproportionately creating more noise pollution for lower income communities.
+From that, we determined if there was a relationship between the permits and whether the community was a minority or low-income.
+Furthermore, we looked at other factors, in particular, the type of permits given when determining the relationship.
+Overall, we determined that there was no evidence that lower-income communities or minority communities were more likely to have more noise exemptions.
+On the contrary, we found moderate evidence for the opposite: higher income communities and communities with less minorities had more noise exemptions, not matter the type of exemptions that were given.
+This is a critical finding because it provides some evidence that noise exemptions are not disproportionately creating more noise pollution for lower income and minority communities.
 
-In terms of the paper's structure, it is organized into three main sections not including the introduction and table of Contents.
-The section # explores the data, its origins, potential issues, and relationships, while the section # provides a deeper discussion of the findings from the data.
-Finally, the section # addresses the paper's limitations and suggests potential next steps for future research.
+In terms of the paper's structure, it is organized into four main sections not including the introduction and table of Contents.
+The section @sec-data explores the data, its origin, potential issues with the data, and how it was cleaned into the data used in this paper.
+Section @sec-results explores potential relationship between certain data variables while section @sec-discussion provides a deeper discussion of the findings from the data including limitations of the study and next steps.
 
@@
 
-All data for this investigation was provided by @OpenDataToronto.
-This data sets after @sec-datacleaning came to 1227 noise exemption permit details.
-Furthermore, this @OpenDataToronto library is created and maintained by government of Toronto.
-Specifically, this data was obtained from three different data sets provided in the library.
+All data for this investigation was provided by Open Data Toronto [@opendatatoronto].
+Open Data Toronto [@opendatatoronto] is a library that is created and maintained by government of Toronto.
+Specifically, this data was obtained from two different data sets provided in the library.
 The first data set was obtained from the Noise Exemption Permits package.
 This data set included everything related to noise exemption permits, including, the day the permit was given, the permit type, the conditions behind it's use, the ward the permit is operating in and more.
-More about this in @tbl-fulldatasample.
+The data was originally taken from the all of the accepted noise exemption permit applications.
 This data set was last updated 2024-09-23.
 
-The rest of the data sets come from the Ward Profiles (25-Ward Model).
-This package contains all the information someone would need about Toronto ward's.
+The rest of the data came from the Ward Profiles (25-Ward Model).
 It was last updated 2024-09-17.
-As for the data sets, the more important one comes from the 2021 Census.
-It contains demographic and socioeconomic data about all of the wards in Toronto.
-It's important to note here that some of the data in that dataset was created based on 25% samples of the census.
-Consequently, variable that were created based on 25% census samples will be denoted in @tbl-fulldatasample The final data set is simply a mapping of the each ward number to the ward's total area in squared kilometers.
+The data used in the packages comes from the 2021 Census and contains demographic and socioeconomic data about all of the wards in Toronto.
+It is important to note here that some of the data in that data set was created based on 25% samples of the census.
+Consequently, variables that were created based on 25% census samples will be denoted in @tbl-fulldatasample
 
-All data analysis was performed using @citeR, @tidyverse, @janitor, and @ggplot2. Also, @kintir was used in order to finalize and create the output of this report.
+All data analysis was performed using R [@citeR], tidyverse [@tidyverse], janitor [@janitor], ggplot2 [@ggplot2], and scales [@scales].
+Also, knitr [@knitr] was used in order to finalize and create the output of this report.
 
@@
 The main reason this occurred was because quite a few of noise exemption permit entries did not have ward numbers.
-Also all permit issue date and permit expected end date values were all converted to years for simplicity purposes as the paper is not concerned about exactly when these permits were issued and ended.
-It's is only concerned about when these permits were issued to determine whether the data is relevant to present day Toronto.
-Additionally, all three of data sets were merged into one data set so each permit has all the information is needs about the ward it is residing in.
-This was soley done to make creating graphs, tables, and other forms of analysis easier.
+Additionally, all permit issue date and permit expected end date values were all converted to years for simplicity purposes as the paper is not concerned about exactly when these permits were issued and ended.
+We also decided to shrink down the number of permit type categories.
+In the end, we decided on three categories, "construction", "non-construction", and "other".
+We did this because we realized that most of the permit categories were construction related categories and those should be treated similarly since many of those permits are critical to maintaining infrasture or building up communities.
+Hence, those exemptions should be treated differently than "non-construction" exemptions, since those do not directly help a community.
+For more information go to @sec-typemerge.
 
@@
 #| label: tbl-fulldatasample
-#| tbl-cap: "Sample of Noise Exmption Dataset"
+#| tbl-cap: "Example Dataset"
 #| echo: false
@@
                           "Expected Permit End Year",
-                          "Ward Total Area (km squared)",
-                          "Ward Total Population", "Ward Median Income (CAD)")
+                          "Ward Total Population", "Ward Average Income (CAD)",
+                          "Ward Minority Percentage")
 
@@
 
-| Variable Name           | Description                                                   |
-|-------------------------|---------------------------------------------------------------|
-| Permit Type             | The type of the permit. More on this in @sec-permittypes       |
-|-------------------------|---------------------------------------------------------------|
-| Ward Number             | The number of the ward of the address the permit's company is working in  |
-|-------------------------|---------------------------------------------------------------|
-| Exemption Issue Year     | The year the permit was issued                                |
-|-------------------------|---------------------------------------------------------------|
-| Expected Permit End Year | The year the government expects this permit to end            |
-|-------------------------|---------------------------------------------------------------|
-| Ward Total Area          | The total area the ward takes in squared kilometers           |
-|-------------------------|---------------------------------------------------------------|
-| Ward Population          | The number of people that live in the ward                   |
-|-------------------------|---------------------------------------------------------------|
-| Ward Median Income       | The median income of people in that ward (CAD)               |
+### Variable Description
+
+**Permit Type**: The type of the permit. Permits can be either Construction or Non-Construction based
+
+**Ward Number**: The number of the ward of the address the permit's company is working in
+
+**Exemption Issue Year**: The year the permit was issued
+
+**Expected Permit End Year**: The year the government expects this permit to end
+
+**Ward Population**: The number of people that live in the ward
+
+**Ward Average Income**: The average income of people in that ward (CAD). This data was taken from a 25% sample of the census.
+
+**Ward Minority Percentage**: Percentage of a ward's population that is a minority
 
@@
 #| label: fig-permityears
-#| tbl-cap: "TODO"
-#| tbl-sub: ["TODO", "TODO"]
-#| layout-ncol: 2
+#| tbl-cap: "Number of Exemptions Per Issue Year and Expected End Year of Permit"
 #| echo: false
@@
 
-noise_exemp_ward_data_og %>% group_by(issue_year) %>% 
-  summarise(n = n_distinct(id)) %>% kable(
-    col.names = c("Issue Year", "Number of Exemptions")
-  )
+issue_year_data <- noise_exemp_ward_data_og %>% group_by(issue_year) %>% 
+  summarise(n = n_distinct(id))
 
-noise_exemp_ward_data_og %>% group_by(expected_end_year) %>% 
-  summarise(n = n_distinct(id)) %>% kable(
-    col.names = c("Expected End Year", "Number of Exemptions")
+end_year_data <- noise_exemp_ward_data_og %>% group_by(expected_end_year) %>% 
+  summarise(n = n_distinct(id))
+
+all_year_data <- issue_year_data %>% 
+  full_join(end_year_data, by=c("issue_year" = "expected_end_year")) %>% 
+  rename(year = 1, issue_year = 2, expected_end_year = 3) %>% 
+  mutate(issue_year = if_else(is.na(issue_year), 0, issue_year), 
+  expected_end_year = if_else(is.na(expected_end_year), 0, expected_end_year))
+
+all_year_data %>% kable(
+    col.names = c("Year", "Exemptions Issued", "Exemptions with Expected End Year")
   )
@@
 
-### Types of Permits {#sec-permittypes}
+# Results {#sec-results}
+
+## Exemption Distribution by Ward
+
+```{r}
+#| label: fig-wardtotalexemptions
+#| fig-cap: "Noise Exemptions Totals by Ward"
+#| fig-subcap: ["Types of Permits By Ward", 
+#| "Boxplot of Total Exemptions Per Ward"]
+#| layout-ncol: 2
+#| out-width: "95%"
+#| fig-align: "center"
+#| echo: false
+#| warning: false
+#| message: false
+
+noise_exemp_ward_data = noise_exemp_ward_data_og %>% 
+  mutate(permit_type = case_when(
+    str_detect(permit_type, "amplified") ~ "non-construction",
+    str_detect(permit_type, "(construction|continuous|crane)") ~ "construction",
+    str_detect(permit_type, "other") ~ "other",
+    TRUE ~ "other"
+  ))
+
+ward_exemptions_bar <- noise_exemp_ward_data %>% 
+  ggplot(aes(x = ward, fill = permit_type)) +
+  geom_bar() +
+  theme_linedraw() +
+  labs(x = "Ward Number", y = "Number of Noise Exemptions")
+
+# this data will primary used for determing relationships, but here
+# it is used to create the box plot
+correlation_data <- noise_exemp_ward_data %>% 
+  group_by(ward, 
+           ward_population, ward_avg_income, ward_minority_percentage) %>%
+  summarise(
+    num_exemptions = n_distinct(id),
+    construction_exemptions = sum(permit_type == "construction"),
+    non_construction_exemptions = sum(permit_type == "non-construction"),
+    other_exemptions = sum(permit_type == "other")
+  )
+
+ward_exemptions_box <- correlation_data %>% 
+  ggplot(aes(x = ward, y = num_exemptions)) +
+  geom_boxplot() +
+  theme_linedraw() +
+  labs(x = "Ward Number", y = "Number of Noise Exemptions") +
+  theme_linedraw() + coord_flip()
+
+ward_exemptions_bar
+ward_exemptions_box
+```
+Based on the @fig-wardtotalexemptions we can see that most wards give out more non-construction permits that construction permits.
+Hence, it's important to see how income is related to number of exemptions based on each type, so general data results are not biased towards non-constructive exemptions.
+
+Moreover, from @fig-wardtotalexemptions and @tbl-wardexemptionssummary, it's clear that wards 13, 11, and especially 10 are outliers in terms of number of exemptions given compared to the rest of the data.
+However, for the purpose of this paper, we will not consider these data points as outliers, so they will not be discarded from the analysis.
+These are still wards of Toronto and their wards cannot be excluded when determining any conclusion or relationships.
+
+## Relationships {#sec-relationships}
+
+### Income Noise Exemption Relationship {#sec-incomeexceptions}
+
+```{r}
+#| label: fig-wardtotalexemptionsscatterplots
+#| fig-cap: "Noise Exemptions by Ward Average Income (CAD)"
+#| fig-subcap: ["Total Exemptions", "Non-construction Exemptions", 
+#| "Construction Exemptions"]
+#| layout-ncol: 3
+#| out-width: "95%"
+#| fig-align: "center"
+#| echo: false
+#| warning: false
+#| message: false
+
+income_num_exemption <- correlation_data %>%  
+  ggplot(aes(ward_avg_income, num_exemptions)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(x = "Average Income", y = "Number of Exemptions") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2) +
+  geom_smooth(method = "lm", se = FALSE) + 
+  scale_x_continuous(labels = label_number())
+
+income_nonconstruction_exemptions <- correlation_data %>% 
+  ggplot(aes(ward_avg_income, non_construction_exemptions)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(x = "Average Income", y = "Number of Non-construction Exemptions") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2) +
+  geom_smooth(method = "lm", se = FALSE) + 
+  scale_x_continuous(labels = label_number())
+
+income_construction_exemptions <- correlation_data %>% 
+  ggplot(aes(ward_avg_income, construction_exemptions)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(x = "Average Income", y = "Number of Construction Exemptions") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2) +
+  geom_smooth(method = "lm", se = FALSE) + 
+  scale_x_continuous(labels = label_number())
+
+income_num_exemption
+income_nonconstruction_exemptions
+income_construction_exemptions
+
+```
+@fig-wardtotalexemptionsscatterplots explore the relationship between the average income of a ward and the number of noise permits given in total and by type.
+Based on the graphs, there seems to be a moderate positive relationship between average income and number of exemptions for all types of permit.
+In this paper, the term moderate description how accurately you can predict a variable with another variable, not how much impact one variable has on the other variable.
+
+Two other things are also important to note here.
+One is that, the income to construction graph seems to have two different postive relationships.
+Additionally, if we hypothetically removed the outliers (ward 10, 11, 13), the relationship would appear to be a bit stronger, particularly in the total exemptions and non-construction exemption cases.
+Plots with outliers excluded are in @fig-wardtotalexemptionsnooutliers.
+
+### Minoirty Relationship
+
+```{r}
+#| label: fig-exemptionsminorityscatterplots
+#| fig-cap: "Noise Exemptions by Ward Minority Percentage"
+#| fig-subcap: ["Total Exemptions", "Non-construction Exemptions", 
+#| "Construction Exemptions"]
+#| layout-ncol: 3
+#| out-width: "95%"
+#| fig-align: "center"
+#| echo: false
+#| warning: false
+#| message: false
+
+minority_num_exemption <- correlation_data %>% 
+  ggplot(aes(ward_minority_percentage, num_exemptions)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(x = "Ward Minority Population (%)", y = "Number of Exemptions") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)  +
+  geom_smooth(method = "lm", se = FALSE)
+
+minority_nonconstruction_exemptions <- correlation_data %>% 
+  ggplot(aes(ward_minority_percentage, non_construction_exemptions)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(x = "Ward Minority Population (%)", y = "Number of Non-construction Exemptions") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)  +
+  geom_smooth(method = "lm", se = FALSE)
+
+minority_construction_exemptions <- correlation_data %>% 
+  ggplot(aes(ward_minority_percentage, construction_exemptions)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(x = "Ward Minority Population (%)", y = "Number of Construction Exemptions") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)  +
+  geom_smooth(method = "lm", se = FALSE)
+
+minority_num_exemption
+minority_nonconstruction_exemptions
+minority_construction_exemptions
+```
+@fig-exemptionsminorityscatterplots explore the relationship between the percentage of minorities in a ward and the number of noise permits given in total and by type.
+Based on the graphs, there does seem to be a negative relationship between the number of exemptions given @sec-incomeexceptions.
+However, this relationship is a bit weaker the one found in @sec-incomeexceptions.
+Similarly, there does seem to be a two different sets of relationships for minority population percentage and its ability to predict construction exemptions.
+Furthermore, like @sec-incomeexceptions, if we consider the case where the potential outliers (ward 10, 11, and 13) are removed, the minority population percentage will look like much stronger predictors for the total number of exemptions and non-construction exemptions.
+Plots with outliers excluded are in @fig-exemptionsminoritynooutliers.
+
+### Average Income, Minority Population, Ward Population Relationship
+
+```{r}
+#| label: fig-minorityincomeconfound
+#| fig-cap: "Exploring Ward Minority Percentage, Average Income, Population Relationship"
+#| fig-subcap: ["Minority v. Income", "Minority v. Population", 
+#| "Income v Population"]
+#| layout-ncol: 3
+#| fig-align: "center"
+#| echo: false
+#| warning: false
+#| message: false
+
+minority_percentage_avg_income <- correlation_data %>% 
+  ggplot(aes(ward_minority_percentage, ward_avg_income)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(x = "Ward Minority Population (%)", y = "Average Ward Income") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2) +
+  geom_smooth(method = "lm", se = FALSE)
+
+minority_population_ward_population <- correlation_data %>% 
+  ggplot(aes(ward_population, ward_minority_percentage)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(y = "Ward Minority Population (%)", x = "Ward Population") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2) +
+  geom_smooth(method = "lm", se = FALSE)
+
+ward_population_avg_income <- correlation_data %>% 
+  ggplot(aes(ward_population, ward_avg_income)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(y = "Ward Average Income", x = "Ward Population") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2) +
+  geom_smooth(method = "lm", se = FALSE) + 
+  scale_y_continuous(labels = label_number())
+
+minority_percentage_avg_income
+minority_population_ward_population
+ward_population_avg_income
+```
+
+@fig-minorityincomeconfound explores the relationship between the percentage of a ward's popualtion that is minority and the ward's average income, the population of a ward and the percentage of that population that is minority, and the ward's average income compared to it's population.
+
+Based on the figures there is little to no evidence that there is a relationship between ward population and the percentage of population that is minority and the same holds true for ward population and ward average income.
+However, there seems to be a very strong relationship between the percentage of a ward's population that is a minority and the ward's average income, indicating that the variables are very good predictors of each other.
+
+\newpage
+
+# Discussion {#sec-discussion}
+
+In this paper, we explored whether or not low-income and minority communities are more likely to be give noise exemption permits compared to other more affluent communities.
+We did this grouping noise permit exemption application data by ward and using public census data to associate each year with its average income and percentage of its population that are minorities.
+This question is relevant since there are numerous research articles which suggest that low-income and minority communities are more affected by noise pollution.
+Thus, we wanted to make sure that Toronto's noise permit exemptions are not further exacerbating this problem.
+Overall, we found no evidence that lower-income communities are given more noise exemptions.
+In fact, we found evidence for opposite: more affluent communities with less minorities had more noise exemptions given.
+
+## Null Findings
+
+As previously mentioned, we found no evidence that minority communities or lower-income communities have more noise-exemptions than higher income communities.
+We determined this because @fig-wardtotalexemptionsscatterplots plots do not hint at any negative relationship between income and noise permit exemptions.
+This was true across both construction and non-construction permits.
+Moreover, we came to same conclusion for minority communities and noise permit exemptions based on @fig-exemptionsminorityscatterplots.
+
+These findings are important for government officials because lower-income and minority communities are typically more affected by noise pollution.
+However, Toronto's noise exemption permits do not contribute to this issue.
+Therefore, city staff and officials should prioritize addressing other sources of noise pollution in these vulnerable areas.
+
+## Unexpected Relationship
+
+Surprisingly, we discovered that typically the higher the ward's average income and the more noise permit exemptions have been accepted.
+This can be seen in @fig-wardtotalexemptionsscatterplots.
+Moreover, based on @fig-wardtotalexemptionsscatterplots we also know that this same relationship exists even after casing for construction and non-construction based permits.
+Furthermore, it is important to note here that for the construction permit graph that there seems to two different potential trend lines.
+This suggests that there might be another variable in addition to a ward's average income that combined could create an extremely accurate prediction of a ward's total number of construction type permits.
+
+While these conclusion might seem strange at first, they make sense practically as well, especially for construction.
+This is because it is reasonable to assume that higher income-areas have more construction permits since more companies will want to invest in those areas, build building.
+Additionally, those same companies and people will need to continue to maintaining those building.
+Thus, these findings can be used by city officials to help raise create initiatives to invest and develop lower income areas.
+
+As for minority communities and noise permit exemptions, we can come to the similar conclusions to similar conclusion to the relationship of ward average income and noise permit exemptions, but in the opposite direction.
+Based on @fig-exemptionsminorityscatterplots, there is some evidence to suggest that having a higher percentage of minorities in a community is associated with less noise permit applications.
+However, this relationship between the two variable in mostly likely indirect.
+More on that in @sec-confounding.
+
+## Confounding Variable Analysis and Potential Casuality {#sec-confounding}
+
+The strongest evidence of a confounding relationship is shown in @fig-minorityincomeconfound, particularly regarding the connection between a ward's average income and the percentage of minorities in that ward.
+A confounding relationship means that while it may seem like there is a direct link between variable X (percentage of minorities) and variable Y (total number of noise exemption permits), the true relationship is influenced by variable Z.
+In this case, variable Z affects both variable X and variable Y, which can create a misleading impression of their relationship.
+In this instance, we suspect that a ward's average income is the confounding variable (Z).
+This is because in there @fig-minorityincomeconfound appears to be a very strong relationship where the higher the percentage of ward's population that is a minority and the lower the ward's average income is.
+Importantly, the average ward income shows a slightly stronger relationship with total noise exemptions than the relationship between noise exemptions and the percentage of minorities.
+This supports the idea that average ward income is the confounding variable, rather than the minority population.
+
+In terms of causality, one of the methods we used to determine this was to see if there was a relationship between a ward's population, its average income, and the percentage of the population that are minorities.
+We did this because it is reasonable to expect that ward's that have higher populations will have more noise exemptions that others.
+However, based on @fig-minorityincomeconfound there is no evidence of a relationship between a ward's population and the other variables of interest.
+This further suggests that average ward income could have a casual effect on the number of noise exemptions a ward has.
+Nonetheless, there are numerous other potential confounds that would need to be tested for in order to come to that conclusion.
+More on confounds in @sec-weaknessnextsteps.
+
+## Weaknesses and Next Steps {#sec-weaknessnextsteps}
+
+The biggest weakness in this study was the use of wards as aggregators.
+The loss of specificity of the data when using wards made a few outliers (Wards 10, 11, and 13) have significant impact on the potential relationships that could be formed with income and noise exemptions.
+Unfortunatly, wards 10, 11, and 13 cannot be ignored until most outliers.
+This is because there are only 25 wards in Toronto, so removing three data points is removing a lot of data.
+Moreover though, the data from these wards are not junk, and they are accurate reflections how noise, income, and minority populations are related in Toronto.
+
+Additionally, using wards to determine socio-economic status can quickly lead to over generalized results.
+For instance, simply because a ward on average is richer than other wards, it does not mean that particular area where the noise permit has been used given is a rich area.
+Hence in future explorations of this topic, the first step would be to determine even more specifically the socioeconomic status of an area.
+This could be based on a specific addresses given in the permit application.
+Then based on this do similar analysis to determine if there is a true relationship between income and noise permits.
+Furthermore, in future studies would also benefit from trying to model the relationship income and noise exemptions.
+Not only does create a more in depth quantification of the relationship, it also provides a much more throughout way of factoring for confounding variables.
+
+\newpage
+
+\appendix
+
+# Appendix {-}
+
+## More on Permit Type Merging {#sec-typemerge}
 
@@
 
-# Amplfied sound is for things like concerts or events
-
-# Construction - self explanatory
-# Continous pour (like pouring concerete, essentially like constructions)
-# Large Carne (need carne to move things, similar to construction)
-# Other sound - self explanatory
-
 full_permit_type_table
@@
 Noise permits are applied to for a variety of reasons.
-These are all the ones listed in the data set.
-In reality there are more and take a look @NoiseToronto if interested in finding out more.
-It's clear that amplified sounds which are often required for things like concerted are the most common form of noise permit exemption.
+These are all the ones that exist in the data set.
+In reality there are more and take a look at Toronto Noise Bylaws [@torontonoise] if interested in finding out more.
+It is important to note that some sound types had subcategories like "amplified sound (level 1)".
+These subcategories follow the same rules as their real category (in this case it is "amplified sound") except they are given a sound level.
 
-TODO: add name and captions
+Types were merged based on if they were relevant to construction or urban maintenance.
+So we had one category called "construction".
+This includes "construction", "continuous pour", "large crane", and "continuous pour and/or large crane" from the original data set.
+The "non-construction" data set includes all of the amplified sound exemptions, and the rest were placed in an "other" type.
 
-| Permit Type           | Description                                                   |
-|-------------------------|---------------------------------------------------------------|
-| Amplified Sound         | Sound made by an electric device. Often required for musical festivals.|
-|-------------------------|---------------------------------------------------------------|
-| Construction             | Self explanatory (TODO)  |
-|-------------------------|---------------------------------------------------------------|
-| Continuous Pour     | Continuously pouring some liquid, like concrete, that cannot be interrupted|
-|-------------------------|---------------------------------------------------------------|
-| Large Crane | Some sort of work that requires a crane (usually for a high rise building)|
-|-------------------------|---------------------------------------------------------------|
-| Other Sound          | Self-explanatory           |
-|-------------------------|---------------------------------------------------------------|
+### Description of Types of Permits
 
-TODO (will add the rest of the types later).
+**Amplified Sound**: Sound made by an electric device. Often required for musical festivals or other public events.
 
-However, for this analysis we will group the permit types into construction related, non-constructed related, and other.
-This is done not only to make things simpler, but because often times these construction related permit types are necessary and cannot be avoid and could potentially benefit a community while things like amplified sound are usually not necessary.
+**Construction**: Sound made to create something, usually buildings.
 
-Non-construction type includes: Amplified
+**Continuous Pour**: Continuously pouring some liquid, almost always concrete, that cannot be interrupted.
 
-Construction type includes: Construction, Continuous Pour, Large Crane
+**Large Crane**: Some sort of work that requires a crane, usually for a high rise building.
 
-Other includes: Other Sounds
+**Continuous Pour and/or Large Crane**: Either continuous pour and/or a large crane
+
+**Other Sound**: Anything that does not fit the other types.
+
+## Addtional Results Details
 
 ```{r}
-#| label: tbl-relevantexemptioncategories
-#| tbl-cap: "Relevant Exemption Categories"
+#| label: fig-wardtotalexemptionsnooutliers
+#| fig-cap: "Noise Exemptions by Ward Average Income (CAD) - No Outliers"
+#| fig-subcap: ["Total Exemptions", "Non-construction Exemptions", 
+#| "Construction Exemptions"]
+#| layout-ncol: 3
+#| out-width: "95%"
+#| fig-align: "center"
 #| echo: false
@@
 
-# Will merge these into construction, non-construction, other
-noise_exemp_ward_data = noise_exemp_ward_data_og %>% 
-  mutate(permit_type = case_when(
-    str_detect(permit_type, "amplified") ~ "non-construction",
-    str_detect(permit_type, "(construction|continuous|crane)") ~ "construction",
-    str_detect(permit_type, "other") ~ "other",
-    TRUE ~ "other"
-  ))
+correlation_data_no_outliers <- correlation_data %>% 
+  filter(!(ward %in% c(10, 11, 13)))
 
-construction_permit_type_table = noise_exemp_ward_data %>% 
-  group_by(permit_type) %>% 
-  summarise(freq = n_distinct(id)) %>% 
-  kable(col.names = c("Type", "Number of Permits"))
+income_num_exemption <- correlation_data_no_outliers %>% 
+  ggplot(aes(ward_avg_income, num_exemptions)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(x = "Average Income", y = "Number of Exemptions") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2) +
+  geom_smooth(method = "lm", se = FALSE) + 
+  scale_x_continuous(labels = label_number())
 
-construction_permit_type_table
+income_nonconstruction_exemptions <- correlation_data_no_outliers %>% 
+  ggplot(aes(ward_avg_income, non_construction_exemptions)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(x = "Average Income", y = "Number of Non-construction Exemptions") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2) +
+  geom_smooth(method = "lm", se = FALSE) + 
+  scale_x_continuous(labels = label_number())
+
+income_construction_exemptions <- correlation_data_no_outliers %>% 
+  ggplot(aes(ward_avg_income, construction_exemptions)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(x = "Average Income", y = "Number of Construction Exemptions") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2) +
+  geom_smooth(method = "lm", se = FALSE) + 
+  scale_x_continuous(labels = label_number())
+
+income_num_exemption
+income_nonconstruction_exemptions
+income_construction_exemptions
 
@@
 
-Even after merging the construction types together. Non-construction or amplified sound types are by far the most common type permit, almost taking up 75% of permits.
-
-## Exemption Distribution by Ward
-
 ```{r}
-#| label: fig-wardtotalexemptions
-#| fig-cap: "Noise Exemptions by Ward"
+#| label: fig-exemptionsminoritynooutliers
+#| fig-cap: "Noise Exemptions by Ward Minority Percentage - No Outliers"
+#| fig-subcap: ["Total Exemptions", "Non-construction Exemptions", 
+#| "Construction Exemptions"]
+#| layout-ncol: 3
+#| out-width: "95%"
+#| fig-align: "center"
 #| echo: false
@@
 
-ward_exemptions_bar <- noise_exemp_ward_data %>% 
-  ggplot(aes(x = ward, fill = permit_type)) +
-  geom_bar() +
+minority_num_exemption <- correlation_data_no_outliers %>% 
+  ggplot(aes(ward_minority_percentage, num_exemptions)) +
+  geom_point(alpha = 0.7) +
   theme_linedraw() +
-  labs(x = "Ward Number", y = "Number of Noise Exemptions")
+  labs(x = "Ward Minority Population (%)", y = "Number of Exemptions") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)  +
+  geom_smooth(method = "lm", se = FALSE)
 
-# TODO: add summary statistics for this graph
+minority_nonconstruction_exemptions <- correlation_data_no_outliers %>% 
+  ggplot(aes(ward_minority_percentage, non_construction_exemptions)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(x = "Ward Minority Population (%)", y = "Number of Non-construction Exemptions") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)  +
+  geom_smooth(method = "lm", se = FALSE)
 
-ward_exemptions_bar
+minority_construction_exemptions <- correlation_data_no_outliers %>% 
+  ggplot(aes(ward_minority_percentage, construction_exemptions)) +
+  geom_point(alpha = 0.7) +
+  theme_linedraw() +
+  labs(x = "Ward Minority Population (%)", y = "Number of Construction Exemptions") + 
+  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)  +
+  geom_smooth(method = "lm", se = FALSE)
+
+minority_num_exemption
+minority_nonconstruction_exemptions
+minority_construction_exemptions
 ```
@@
 #| label: tbl-wardexemptionssummary
-#| tab-cap: "Noise Exemptions Summary And Outliers"
-#| tbl-subcap: ["Summary", "Outliers"]
-#| layout-ncol: 2
+#| tbl-cap: "Summary of Total Noise Exemptions By Ward"
 #| echo: false
@@
     max = max(exemptions),
-    IQR = IQR(exemptions)
+    IQR = IQR(exemptions),
+    mean = mean(exemptions),
+    sd = sd(exemptions)
   )
 
-# calculate outliers
-# TODO: add titles
+# ward 10, 11, 13, but this table doesn't need to be shown just described
 ward_exemption_outliers <- ward_exemptions %>% filter(
@@
 
-five_number_summary %>% kable()
-ward_exemption_outliers %>% kable()
-```
-Based on the @fig-wardtotalexemptions we can see that most wards give out more non-construction permits that construction permits.
-Hence, it's important to see how income is related to number of exemptions based on each type, so general data results are not biased towards non-constructive exemptions.
-
-Moreover, from @fig-wardtotalexemptions and @tbl-wardexemptionssummary, it's clear that wards 13, 11, and especially ten are outlines in terms of number of exemptions given compared to the rest of the data.
-However, for the purpose of this paper, we will not consider these data points as outliers, so they will not be discarded from the analysis.
-This is these are still wards of Toronto and their wards cannot be excluded when determining any conclusion or relationships.
-
-## Income Noise Exemption Relationship
-
-```{r}
-#| label: fig-wardtotalexemptionsscatterplots
-#| fig-cap: "Noise Exemptions by Ward"
-#| fig-subcap: ["TODO", "TODO", "TODO"]
-#| layout-ncol: 3
-#| echo: false
-#| warning: false
-#| message: false
-
-# TODO: change this so that the general num_exemptions graph is
-# in it's own code black and the sub groups have a layout-ncol-2
-
-correlation_data <- noise_exemp_ward_data %>%
-  group_by(ward, ward_median_income, ward_area_km2, 
-           ward_population) %>% 
-  summarise(
-    num_exemptions = n_distinct(id),
-    construction_exemptions = sum(permit_type == "construction"),
-    non_construction_exemptions = sum(permit_type == "non-construction"),
-    other_exemptions = sum(permit_type == "other")
+five_number_summary %>% 
+  kable(
+    col.names = c("Minimum", "1st Quartile", "Median", "3rd Quartile", "Maximum", "IQR", "Mean", "Standard Deviation"),
   )
 
-income_num_exemption <- correlation_data %>% 
-  ggplot(aes(ward_median_income, num_exemptions)) +
-  geom_point(alpha = 0.7) +
-  theme_linedraw() +
-  labs(x = "Median Income", y = "Number of Exemptions") + 
-  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)
-
-income_nonconstruction_exemptions <- correlation_data %>% 
-  ggplot(aes(ward_median_income, non_construction_exemptions)) +
-  geom_point(alpha = 0.7) +
-  theme_linedraw() +
-  labs(x = "Median Income", y = "Number of Non-construction Exemptions") + 
-  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)
-
-income_construction_exemptions <- correlation_data %>% 
-  ggplot(aes(ward_median_income, construction_exemptions)) +
-  geom_point(alpha = 0.7) +
-  theme_linedraw() +
-  labs(x = "Median Income", y = "Number of Construction Exemptions") + 
-  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)
-
-income_num_exemption
-income_nonconstruction_exemptions
-income_construction_exemptions
-
-# keep in mind that relationship looks weaker after potetnial outlier point
-# removed
-
-# TODO: ADD a section about these graphs even after they have been removed
 ```
-
-Based on the @fig-wardtotalexemptionsscatterplots.
-There is not seem to be a negative relation ship between median ward income and number of exemptions.
-However, by looking at the plot, there does seem to me a moderate positive relationship between median income and number of exemptions for all types of permit, although the construction permit plot's relationship does appear to be weaker.
-
-It is important to note that if we hypothetically removed the outliers (ward 10, 11, 13), the relationship would appear to be a bit weaker, but there still appears to be a moderate positive relationship.
-
-## Testing for Hidden Confounding relationships
-
-```{r}
-#| label: fig-confoudingplots
-#| fig-cap: "Ward Population impact on Medium Impact"
-#| fig-subcap: ["TODO", "TODO"]
-#| layout-ncol: 2
-#| echo: false
-#| warning: false
-#| message: false
-
-pop_exemptions = correlation_data %>% ggplot(aes(ward_population, num_exemptions)) +
-  geom_point(alpha = 0.7) +
-  theme_linedraw() +
-  labs(y = "Exemptions", x = "Population") + 
-  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)
-
-correlation_data_confound <- noise_exemp_ward_data %>% 
-  select(ward, ward_area_km2, ward_population, ward_median_income)
-
-ward_income_pop <- correlation_data_confound %>% 
-  ggplot(aes(ward_population, ward_median_income)) +
-  geom_point(alpha = 0.7) +
-  theme_linedraw() +
-  labs(y = "Median Income", x = "Population") + 
-  geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)
-
-# ward_income_area <- correlation_data %>% 
-#   ggplot(aes(ward_area_km2, ward_median_income)) +
-#   geom_point(alpha = 0.7) +
-#   theme_linedraw() +
-#   labs(y = "Median Income", x = "ward area (kms squared)") + 
-#   geom_text(aes(label = ward), vjust = 1.5, hjust = 0.5, size = 2)
-
-ward_income_pop
-pop_exemptions
-# ward_income_area
-
-```
-@fig-confoudingplots explores a ward's population is a good predictor of a ward's population and the number of permit exemptions given.
-Based on the plots there does seem to be a moderate positive relationship between a ward's population and a ward's medium income.
-Likewise, there is a moderate positive relationship between ward population and the number of noise exemptions a ward has given although this reltionship is not as strong as population and median income.
-
-However, we need to also consider that if we hypothetically remove the outlines from population graph.
-Then there appeared to be very little if not no relationship between population and median income.
-
-\newpage
-
-# Discussion
-
-## First discussion point {#sec-first-point}
-
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
-
-## Second discussion point
-
-## Third discussion point
-
-## Weaknesses and Next Steps
-
-Weaknesses/Limitations and next steps should also be included.
-
-\newpage
-
-<!-- TODO: DELETE THIS CODE AND MOVE IT SOMEWHERE ELSE  -->
-
-# Special Thanks
-
-@citeR and @rohan.
 

### 2024-09-26 scripts/01-download_data.R

@@ -37,8 +37,6 @@ ward_resources <- list_package_resources("6678e1a6-d25f-4dff-b2b7-aa8f042bc2eb")
 
 # in xlsx
 census_dataset <- get_resource("16a31e1d-b4d9-4cf0-b5b3-2e3937cb4121")
-ward_areas <- get_resource("9398da69-0622-4eb1-a125-4f8c7f1016a4")
-
 
 #### Save data ####
 # change the_raw_data to whatever name you assigned when you downloaded it.
@@ -47,8 +45,6 @@ write_csv(noise_exmeption_dataset,
 # for some reason r wasn't finding the function
 # unless writexl package was explicitly defined
 writexl::write_xlsx(census_dataset, "./data/raw_data/raw_census_dataset.xlsx")
-writexl::write_xlsx(ward_areas, 
-                    "./data/raw_data/raw_ward_areas_dataset.xlsx")
 
 
          

### 2024-09-26 scripts/02-data_cleaning.R

@@ -35,7 +35,6 @@ clean_noise_exmp_data <- raw_noise_exmp_data %>%
   select(!c(issue_date, expected_end_date))
 
 ## cleaning 2021 census data
-
 raw_census_one_variable_data <- 
   read_excel("./data/raw_data/raw_census_dataset.xlsx", 
              sheet = "2021 One Variable") 
@@ -60,63 +59,67 @@ clean_ward_pop_data <- clean_ward_pop_data %>%
   # there are no na's put this code is added for reproduce ability
   drop_na() %>% 
   filter(str_detect(ward, "Ward")) %>% 
-  mutate(ward = as.numeric(str_split_i(ward, " ", i = 2))) %>% 
+  mutate(ward = as.numeric(str_split_i(ward, " ", i = 2)),
+         ward_population = parse_number(ward_population)) %>% 
   select(ward, ward_population)
 
-## getting each ward's low income percentage
-clean_ward_median_income_data <- raw_census_one_variable_data %>% 
+## Getting Average Total Income By Ward
+
+clean_ward_avg_income_data <- raw_census_one_variable_data %>% 
   clean_names()
 
-
-clean_ward_median_income_data <- clean_ward_median_income_data %>% 
+clean_ward_avg_income_data <- clean_ward_avg_income_data %>% 
   filter(x2 == "Toronto" | 
            str_detect(city_of_toronto_profiles,
-                "Median total income in 2020 among recipients")) %>% 
+                      "Average total income in 2020 among recipients")) %>% 
   select(!city_of_toronto_profiles) %>% 
   distinct()
 
 # flipping rows and columns
-clean_ward_median_income_data <- as.data.frame(t(as.matrix(
-  clean_ward_median_income_data)))
+clean_ward_avg_income_data <- as.data.frame(t(as.matrix(
+  clean_ward_avg_income_data)))
 
-clean_ward_median_income_data <- clean_ward_median_income_data %>% 
-  rename(ward = 1, ward_median_income = 2) %>%
+clean_ward_avg_income_data <- clean_ward_avg_income_data %>% 
+  rename(ward = 1, ward_avg_income = 2) %>%
   # there are no na's put this code is added for reproduce ability
   drop_na() %>% 
   filter(str_detect(ward, "Ward")) %>% 
   mutate(ward = as.numeric(str_split_i(ward, " ", i = 2))) %>% 
-  mutate(ward_median_income = parse_number(ward_median_income))
[…]

### 2024-09-26 paper/paper.qmd
proof read text

@@
 This paper investigates if low-income and minority communities are more likely to receive noise permit exemptions.
-Contrary to expectations, the findings indicate that low-income and minority communities do not have a higher likelihood of receiving exemptions, In fact we found evidence for the opposite being true, that communities with higher income and less minorities recieved more noise exemption permits."
+Contrary to expectations, the findings indicate that low-income and minority communities do not have a higher likelihood of receiving exemptions.
+In fact we found evidence for the opposite being true, that communities with higher income and less minorities recieved more noise exemption permits."
 format: pdf
@@
 Noise pollution has been linked with heart issues, irregular sleep, hearing loss, and mental health issues [@IberdrolaNoisePollution].
-Hence, Toronto city officials have identified noise pollution as a problem and created Noise Bylaws [@torontonoise] to limit at what times certain noises like construction can occur.
+Hence, Toronto city officials have identified noise pollution as a problem and created Noise Bylaws [@torontonoise] to limit what times certain noises like construction can occur.
 However, sometimes noise needs to be allowed during quiet hours.
-For instance, some building need to continue pouring cement overnight, otherwise the building will have safety risks.
+For instance, some building may need to continue pouring cement overnight, otherwise the building will have increased safety risks.
 As a result, the Toronto government has also created a system of noise exemptions.
@@
 
-In terms of the paper's structure, it is organized into four main sections not including the introduction and table of Contents.
-The section @sec-data explores the data, its origin, potential issues with the data, and how it was cleaned into the data used in this paper.
-Section @sec-results explores potential relationship between certain data variables while section @sec-discussion provides a deeper discussion of the findings from the data including limitations of the study and next steps.
+In terms of the paper's structure, it is organized into four main sections not including the introduction.
+The section @sec-data explores the data, its origin, and how it was cleaned.
+Section @sec-results explores potential relationships between certain data variables.
+Finally, section @sec-discussion provides a deeper discussion of the findings from the data including limitations of the study and next steps.
 
@@
 The first data set was obtained from the Noise Exemption Permits package.
-This data set included everything related to noise exemption permits, including, the day the permit was given, the permit type, the conditions behind it's use, the ward the permit is operating in and more.
-The data was originally taken from the all of the accepted noise exemption permit applications.
+This data set included everything related to noise exemption permits, including, the day the permit was given, the permit type, the ward the permit is operating in and more.
+The data was originally taken from all of the accepted noise exemption permit applications.
 This data set was last updated 2024-09-23.
 
-The rest of the data came from the Ward Profiles (25-Ward Model).
+The rest of the data came from the Ward Profiles (25-Ward Model) package.
 It was last updated 2024-09-17.
@@
 It is important to note here that some of the data in that data set was created based on 25% samples of the census.
-Consequently, variables that were created based on 25% census samples will be denoted in @tbl-fulldatasample
+Variables that were created based on 25% census samples will be denoted in @tbl-fulldatasample.
 
@@
 
-Some noise exemption data had missing values and those were rows.
-The main reason this occurred was because quite a few of noise exemption permit entries did not have ward numbers.
-Additionally, all permit issue date and permit expected end date values were all converted to years for simplicity purposes as the paper is not concerned about exactly when these permits were issued and ended.
+Some noise exemption data had missing values and those rows were removed.
+The main reason this occurred was because some noise exemption permit entries did not have ward numbers.
+Additionally, all permit issue dates and permit expected end dates were converted to years for simplicity purposes as the paper is not concerned about exactly when these permits were issued and ended.
 We also decided to shrink down the number of permit type categories.
 In the end, we decided on three categories, "construction", "non-construction", and "other".
-We did this because we realized that most of the permit categories were construction related categories and those should be treated similarly since many of those permits are critical to maintaining infrasture or building up communities.
+We did this because we realized that most of the permit categories were construction related and those should be treated similarly since many of those permits are critical to maintaining infrastructure or building up communities.
 Hence, those exemptions should be treated differently than "non-construction" exemptions, since those do not directly help a community.
@@
 
-**Permit Type**: The type of the permit. Permits can be either Construction or Non-Construction based
+**Permit Type**: The type of the permit. Permits can be either type Construction, Non-Construction or Other
 
-**Ward Number**: The number of the ward of the address the permit's company is working in
+**Ward Number**: The ward in which the permit is allowed to be used
 
@@
 
-**Ward Minority Percentage**: Percentage of a ward's population that is a minority
+**Ward Minority Percentage**: Percentage of a ward's population that is a minority. This data was taken from a 25% sample of the census.
 
@@
 
-The table shows that almost all permits have been issued in 2024 and expected to end sometime in 2024 or 2025.
-It's is important to note that this was not deliberate choice and that this is merely a limitation of the data.
+The table shows that almost all permits have been issued in 2024 and are expected to end sometime in 2024 or 2025.
+It is important to note that this was not deliberate choice and that this is merely a limitation of the data.
 However, this should not be viewed as a limitation because having all of the data being from 2019 and beyond makes sure that all this data is relevant to how noise exemption permits are issued now.
@@
 ```
-Based on the @fig-wardtotalexemptions we can see that most wards give out more non-construction permits that construction permits.
-Hence, it's important to see how income is related to number of exemptions based on each type, so general data results are not biased towards non-constructive exemptions.
+Based on the @fig-wardtotalexemptions we can see that most wards give out more non-construction permits than construction permits.
+Hence, it will be important to see how income and minority population are related to number of exemptions based on each type, so general data results are not biased towards non-constructive exemptions.
 
-Moreover, from @fig-wardtotalexemptions and @tbl-wardexemptionssummary, it's clear that wards 13, 11, and especially 10 are outliers in terms of number of exemptions given compared to the rest of the data.
+Moreover, from @fig-wardtotalexemptions and @tbl-wardexemptionssummary, it is clear that wards 13, 11, and especially 10 are outliers in terms of number of exemptions given compared to the rest of the data.
 However, for the purpose of this paper, we will not consider these data points as outliers, so they will not be discarded from the analysis.
@@
 @fig-wardtotalexemptionsscatterplots explore the relationship between the average income of a ward and the number of noise permits given in total and by type.
-Based on the graphs, there seems to be a moderate positive relationship between average income and number of exemptions for all types of permit.
-In this paper, the term moderate description how accurately you can predict a variable with another variable, not how much impact one variable has on the other variable.
+Based on the graphs, there seems to be a moderate positive relationship between average income and the number of exemptions for all types of permit.
+In this paper, the term moderate describes how accurately you can predict a variable with another variable, not how much impact one variable has on the other variable.
 
 Two other things are also important to note here.
-One is that, the income to construction graph seems to have two different postive relationships.
+One is that, the income to construction graph seems to have two different positive relationships.
 Additionally, if we hypothetically removed the outliers (ward 10, 11, 13), the relationship would appear to be a bit stronger, particularly in the total exemptions and non-construction exemption cases.
@@
 However, this relationship is a bit weaker the one found in @sec-incomeexceptions.
-Similarly, there does seem to be a two different sets of relationships for minority population percentage and its ability to predict construction exemptions.
-Furthermore, like @sec-incomeexceptions, if we consider the case where the potential outliers (ward 10, 11, and 13) are removed, the minority population percentage will look like much stronger predictors for the total number of exemptions and non-construction exemptions.
+Similarly, there does seem to be two different sets of relationships for minority population percentage and its ability to predict construction exemptions.
+Furthermore, like @sec-incomeexceptions, if we consider the case where the potential outliers (ward 10, 11, and 13) are removed, the minority population percentage will look a bit better as predictors for the total number of exemptions and non-construction exemptions.
 Plots with outliers excluded are in @fig-exemptionsminoritynooutliers.
@@
 
-@fig-minorityincomeconfound explores the relationship between the percentage of a ward's popualtion that is minority and the ward's average income, the population of a ward and the percentage of that population that is minority, and the ward's average income compared to it's population.
+@fig-minorityincomeconfound explores the relationship between the percentage of a ward's population that is a minority and the ward's average income; the population of a ward and the percentage of that population that is minority; and the ward's average income compared to it's population.
 
@@
 In this paper, we explored whether or not low-income and minority communities are more likely to be give noise exemption permits compared to other more affluent communities.
-We did this grouping noise permit exemption application data by ward and using public census data to associate each year with its average income and percentage of its population that are minorities.
+We did this by grouping noise permit exemption application data by ward and using public census data to associate each ward with its average income and percentage of its population that are minorities.
 This question is relevant since there are numerous research articles which suggest that low-income and minority communities are more affected by noise pollution.
 Thus, we wanted to make sure that Toronto's noise permit exemptions are not further exacerbating this problem.
-Overall, we found no evidence that lower-income communities are given more noise exemptions.
+Overall, we found no evidence that lower-income and minority communities are given more noise exemptions.
 In fact, we found evidence for opposite: more affluent communities with less minorities had more noise exemptions given.
@@
 
-Surprisingly, we discovered that typically the higher the ward's average income and the more noise permit exemptions have been accepted.
+Surprisingly, we discovered that typically the higher the ward's average income the more noise permit exemptions have been accepted.
 This can be seen in @fig-wardtotalexemptionsscatterplots.
@@
 While these conclusion might seem strange at first, they make sense practically as well, especially for construction.
-This is because it is reasonable to assume that higher income-areas have more construction permits since more companies will want to invest in those areas, build building.
+This is because it is reasonable to assume that higher income-areas have more construction permits since more companies will want to invest in those areas and build building.
 Additionally, those same companies and people will need to continue to maintaining those building.
-Thus, these findings can be used by city officials to help raise create initiatives to invest and develop lower income areas.
+Thus, these findings can be used by city officials to help create initiatives to invest and develop lower income areas.
 
-As for minority communities and noise permit exemptions, we can come to the similar conclusions to similar conclusion to the relationship of ward average income and noise permit exemptions, but in the opposite direction.
+As for minority communities and noise permit exemptions, we can come to similar conclusions for the relationship of ward average income and noise permit exemptions, but in the opposite direction.
 Based on @fig-exemptionsminorityscatterplots, there is some evidence to suggest that having a higher percentage of minorities in a community is associated with less noise permit applications.
@@
 In this instance, we suspect that a ward's average income is the confounding variable (Z).
-This is because in there @fig-minorityincomeconfound appears to be a very strong relationship where the higher the percentage of ward's population that is a minority and the lower the ward's average income is.
+This is because in @fig-minorityincomeconfound there appears to be a very strong relationship where the higher the percentage of ward's population that is a minority, the lower the ward's average income is.
 Importantly, the average ward income shows a slightly stronger relationship with total noise exemptions than the relationship between noise exemptions and the percentage of minorities.
@@
 In terms of causality, one of the methods we used to determine this was to see if there was a relationship between a ward's population, its average income, and the percentage of the population that are minorities.
-We did this because it is reasonable to expect that ward's that have higher populations will have more noise exemptions that others.
+We did this because it is reasonable to expect that the wards with higher populations will have more noise exemptions than others.
 However, based on @fig-minorityincomeconfound there is no evidence of a relationship between a ward's population and the other variables of interest.
@@
 The biggest weakness in this study was the use of wards as aggregators.
-The loss of specificity of the data when using wards made a few outliers (Wards 10, 11, and 13) have significant impact on the potential relationships that could be formed with income and noise exemptions.
-Unfortunatly, wards 10, 11, and 13 cannot be ignored until most outliers.
+The loss of specificity when using wards made a few outliers (Wards 10, 11, and 13) have significant impact on the potential relationships that could be formed with income and noise exemptions.
+Unfortunately, wards 10, 11, and 13 cannot be ignored unlike most outliers.
 This is because there are only 25 wards in Toronto, so removing three data points is removing a lot of data.
-Moreover though, the data from these wards are not junk, and they are accurate reflections how noise, income, and minority populations are related in Toronto.
+Moreover, the data from these wards are not junk, and they are accurate reflections how noise, income, and minority populations are related in Toronto.
 
 Additionally, using wards to determine socio-economic status can quickly lead to over generalized results.
-For instance, simply because a ward on average is richer than other wards, it does not mean that particular area where the noise permit has been used given is a rich area.
+For instance, simply because a ward on average is richer than other wards, it does not mean that particular address where the noise permit has been used given is a rich area.
 Hence in future explorations of this topic, the first step would be to determine even more specifically the socioeconomic status of an area.
@@
 Then based on this do similar analysis to determine if there is a true relationship between income and noise permits.
-Furthermore, in future studies would also benefit from trying to model the relationship income and noise exemptions.
-Not only does create a more in depth quantification of the relationship, it also provides a much more throughout way of factoring for confounding variables.
+Furthermore, future studies would also benefit from trying to model the relationship income and noise exemptions.
+Not only does this create a more in depth quantification of the relationship, it also provides a much more thorough way of factoring for confounding variables.
 
@@
 These are all the ones that exist in the data set.
-In reality there are more and take a look at Toronto Noise Bylaws [@torontonoise] if interested in finding out more.
-It is important to note that some sound types had subcategories like "amplified sound (level 1)".
+In reality there are more, so take a look at Toronto Noise Bylaws [@torontonoise] for more information.
+Also, it is important to note that some sound types had subcategories like "amplified sound (level 1)".
 These subcategories follow the same rules as their real category (in this case it is "amplified sound") except they are given a sound level.
@@
 
-# ward 10, 11, 13, but this table doesn't need to be shown just described
+# ward 10, 11, 13, but this table doesn't need to be shown but only exists
+# for author's reference
 ward_exemption_outliers <- ward_exemptions %>% filter(
